---
title: "Course 8 Week 4 Machine Learning Project"
author: "Xiao"
date: "1/22/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here

## Data 
The training data for this project are available here: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv \
The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv \

Load two datasets
```{r}
train <- read.csv('pml-training.csv', na.strings=c("NA","#DIV/0!",""))
test <- read.csv("pml-testing.csv", na.strings=c("NA","#DIV/0!",""))
dim(train)
```

```{r}
table(train$classe)
```

## Preprocessing
### Partition the training set
Separate training data into a training set and a validation set for cross validation.
```{r}
library(caret)
set.seed(531)
trainset <- createDataPartition(train$classe, p = 0.8, list = FALSE)
training <- train[trainset, ]
validation <- train[-trainset, ]
dim(training)
```

### Feature selection
Clean up near zero variance features, columns with missing values and descriptive fields.
```{r}
# exclude near zero variance features
nzvcol <- nearZeroVar(training)
training <- training[, -nzvcol]

# exclude columns with more than 40% missing values exclude descriptive
# columns like name etc
cntlength <- sapply(training, function(x) {
  sum(!(is.na(x) | x == ""))
  })
nullcol <- names(cntlength[cntlength < 0.6 * length(training$classe)])
#nullcol
descriptcol <- c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", 
    "cvtd_timestamp", "new_window", "num_window")
excludecols <- c(descriptcol, nullcol)
training <- training[, !names(training) %in% excludecols]
training[, 'classe'] <- as.factor(training[, 'classe'])
validation[, 'classe'] <- as.factor(validation[, 'classe'])
dim(training)
```

## Train model
Use random forest as our model as implemented in the randomForest package by Breiman's random forest algorithm (based on Breiman and Cutler's original Fortran code) for classification and regression.
```{r}
library(randomForest)
rfModel <- randomForest(classe ~ ., data = training, importance = TRUE, ntrees = 1)
```

## Model validation
### Training set accuracy
```{r}
ptraining <- predict(rfModel, training)
#head(ptraining, 100)
#head(training$classe, 100)
#install.packages("e1071", dep = TRUE)
library(e1071)
print(confusionMatrix(ptraining, training$classe))
```
The model performs excellent against the training set, but we need to cross validate the performance against the held out set and see if we have avoided overfitting.

### Cross validation set accuracy
See how the model performs on the cross validation set that we held out from training.
```{r}
pvalidation <- predict(rfModel, validation)
print(confusionMatrix(pvalidation, validation$classe))
```
The cross validation accuracy is 99.5% and the out-of-sample error is therefore 0.5% so our model performs good.

## Test set prediction
```{r}
ptest <- predict(rfModel, test)
ptest
```
The logic behind using the random forest method as the predictor rather than other methods or a combination of various methods is:

- Random forests are suitable when to handling a large number of inputs, especially when the interactions between variables are unknown.\
- Random forest’s built in cross-validation component that gives an unbiased estimate of the forest’s out-of-sample (or bag) (OOB) error rate.\
- A Random forest can handle unscaled variables and categorical variables. This is more forgiving with the cleaning of the data.\

Save the output to files according to instructions and post it to the submission page.
```{r}
answers <- as.vector(ptest)

pml_write_files = function(x) {
    n = length(x)
    for (i in 1:n) {
        filename = paste0("problem_id_", i, ".txt")
        write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, 
            col.names = FALSE)
    }
}

pml_write_files(answers)
```








